Linting Linearity in Core/System FC

### 2022-11-11 proposal.tex
Add some comments
Expand letrec solution

\documentclass[10pt, a4paper, draft]{article}

@@

@@

@@

Optimisations can create a letrec which uses a variable linearly. The following
example uses 'x' linearly, but this is not seen by the linter.
\begin{code}

letrec f = \case
        True -> f False
        False -> x
in f True
\end{code}

@@

@@

\begin{itemize}
\item https://gitlab.haskell.org/ghc/ghc/-/issues/20058
\item https://gitlab.haskell.org/ghc/ghc/-/issues/18768
\end{itemize}

@@

>>>>

\documentclass[a4paper, draft]{article}

@@

\section{Background}

\section{Technical Details}

@@

What do we currently do?

When we lint a core expression, we get both its type and its usage environment.
That means that to lint linearity in an expression, whenever we come across a
free variable we compute its usage environment and take it into account

@@

Optimisations can create a letrec which uses a variable linearly. The following
example uses 'x' linearly, but this is not seen by the linter.
\begin{code}
letrec f = \case
        True -> f False
        False -> x
in f True
\end{code}

@@

\textbf{Re-explained:} to compute the usage environment of the recursive
let-bound function $f$ when applied, we compute $Z$ such that for all branch
alternatives $U,V,...$, $U \subset Z$, $V \subset Z$ and so on by tracking the
multiplicities and usage environments of variables that show up in the body and
by emitting a special keyword $rec$ everytime we find a saturated call of $f$
(how to handle unsaturated calls?) (how to have a more general solution that
doesn't require a keyword, perhaps something to do with fixed points); Then we
scale $Z \setminus \{(rec,_)\}$ by $Z[rec]$ and get $T$ which is the usage
environment of $f$ when saturated.

\textbf{Example 1 revisited:}
\begin{enumerate}
    \item Take $U = \{rec := 1\}$
    \item Take $V = \{x := 1\}$
    \item Take $Z = \{x := 1, rec := 1\}$
    \item Take $\pi = Z[rec] = 1$
    \item Take $W = Z \setminus \{rec\} = \{x := 1\}$
    \item Take $T = \pi W = \{x := \pi * 1\} = \{x := 1\}$
    \item Linearity OK
\end{enumerate}

\textbf{Example 2 revisited:}
\begin{enumerate}
    \item Take $U = \{rec := 2\}$
    \item Take $V = \{x := 1\}$
    \item Take $Z = \{x := 1, rec := 1\}$
    \item Take $\pi = Z[rec] = 2$
    \item Take $W = Z \setminus \{rec\} = \{x := 1\}$
    \item Take $T = \pi W = \{x := \pi * 1\} = \{x := 2\}$
    \item Linearity not OK
\end{enumerate}

@@

For case expressions, the usage environment is computed by checking all branches
and taking sup. However, this trick doesn't work when there are no branches.

\begin{itemize}
\item https://gitlab.haskell.org/ghc/ghc/-/issues/20058
\item https://gitlab.haskell.org/ghc/ghc/-/issues/18768

\item (1) Just like a case expression remembers its type (Note [Why does Case have a
'Type' field?] in Core.hs), it should remember its usage environment. This data
should be verified by Lint.

\item (2) Once this is done, we can remove the Bottom usage and the second field of
UsageEnv. In this step, we have to infer the correct usage environment for empty
case in the typechecker.
\end{itemize}

@@

This example is well typed: a function is linear if it consumes its argument
exactly once when it's consumed exactly once. It seems like the function isn't
linear since it won't consume x because of the empty case, however, that also
means f won't be consumed due to the same empty case, thus linearity is
preserved.

\begin{code}
* In the case of empty types (see Note [Bottoming expressions]), say
  data T
we do NOT want to replace
  case (x::T) of Bool {}   -->   error Bool "Inaccessible case"
because x might raise an exception, and *that*'s what we want to see!
(#6067 is an example.) To preserve semantics we'd have to say
   x `seq` error Bool "Inaccessible case"
but the 'seq' is just such a case, so we are back to square 1.
\end{code}

There are three different problems:

\begin{itemize}
\item castBottomExpr converts (case x :: T of {}) :: T to x.
\item Worker/wrapper moves the empty case to a separate binding
\item CorePrep eliminates empty case, just like point 1 (See -- Eliminate empty
    case in GHC.CoreToStg.Prep
\end{itemize}

With castBottomExpr, we get the example above to
\begin{code}
    f = \ @a (x (%1) :: a) -> ()
\end{code}
And if we don't, 
\begin{code}
    f = \ @a (x (%1) :: a) -> case () of {}
\end{code}
And that supposedly if we had a usage environment in the case expression we
could avoid the error. How is it typed without the transformation in face
of the bottom? (Even knowing that theoretically it's because of divergence?)
